### 2018-12-28 Sections/intro_to_QM.tex
Formatting update

\dropcap{T}{he} laws of quantum mechanics are inherently different to those of classical mechanics in virtually every way, and bring with them their own unique notation for:
\begin{itemize}
	\item The representation of quantum states.
	\item The evolution of quantum systems.
	\item The measurement of quantum states.
\end{itemize}
We review all of these formalisms in the upcoming sections, and present the necessary mathematical framework to enable following the derivations presented here.

@@

\subsection{Features of quantum mechanics}\index{Characteristics of quantum mechanics}

Classical and quantum physics are fundamentally different in a number of extremely important ways. These differences are what allow quantum mechanics to enable the entire new field of quantum technologies, most notably quantum computing, which offers massive computational gains over traditional classical computing. The same applies to the novel and groundbreaking promises of the other key quantum technologies, such as quantum cryptography, quantum teleportation, and quantum-enhanced metrology. All of these gains are achieved by exploiting these unique new characteristics offered by quantum systems, that are physically impossible in a classical context.

The key conceptual differences between classical and quantum mechanics are:
\begin{itemize}
	\item Superposition\index{Superpositions}: the ability of quantum systems to be in multiple states simultaneously (not probabilistically!). This arises since quantum states can always be interpreted as waves, which are inherently delocalised and non-particle-like.
	\item Quantum amplitudes: quantum states are not governed by classical probabilities, but by \textit{quantum amplitudes}, which are complex numbers rather than real numbers.\index{Quantum amplitudes}
	\item Entanglement: a type of non-local quantum correlation, stronger than any possible classical correlation.\index{Entanglement}
	\item Unitary evolution: the free evolution of quantum systems is restricted to being described by unitary operators. This implies that all quantum evolution is perfectly time-reversible\index{Time-reversal}, since all unitary operators are invertible.\index{Evolution}
	\item Measurement collapse: the measurement of a quantum system changes its state in general, \textit{collapsing} it to the state corresponding to the measurement outcome\footnote{An open and intuitively confusing conundrum in quantum theory is to resolve the question of what exactly it is that constitutes `measurement' -- the \textit{measurement paradox}\index{Measurement paradox}. Given that all quantum evolution is supposed to be governed by unitary dynamics, whereas measurements are not, at what point in physical reality does an operation transition from one to the other? This is an extremely important open question, particularly from the perspective of quantum foundations and the philosophy of physics. This has led to the emergence of countless theories for the interpretation of quantum mechanics\index{Interpretation of quantum mechanics}, which is an essentially untestable field of research, more akin to fundamentalist religious movements than physical science. However, the resolution is largely unimportant in a practical technological context, where it suffices to define `measurement' as being `when quantum information becomes classical information', e.g once we are able to read it off a measurement device's LCD screen!}.\index{Measurement}
	\item Non-determinism: quantum systems exhibit genuine non-determinism, associated with the randomness of the measurement collapse\index{Measurement!Collapse} process\footnote{When we say `random' we \textit{really} mean random, not `quasi-random but actually deterministic'\index{Quasi-randomness}. This true randomness is another of the unique and surprising features of quantum mechanics. Unlike classical physics, quantum mechanics is inherently and provably non-deterministic! Specifically, it has been proven theoretically \cite{bib:BellNoHiddenVariables} that so-called `hidden variable theories'\index{Hidden variable theories} are incompatible with quantum physics, which rules out any form of quasi-randomness and mandates true randomness.}.\index{Non-determinism}
	\item Exponential complexity: the number of unique parameters required to fully describe a quantum system grows exponentially with its `size' (e.g the number of constituent subsystems, such as quantum particles or qubits). This is in contrast to classical systems whose state complexity grows only linearly with system size, since it's always possible to describe constituent subsystems independently (there is no notion of entanglement). Therefore the number of parameters required to characterise classical systems accumulates additively, whereas for quantum systems it grows multiplicatively.\index{Exponential complexity}
\end{itemize}

@@

Mathematically, we can employ the following alternate state vector representations, where the latter is referred to as \textit{Dirac notation} (with $\ket{n}$ denoting the basis states),
\begin{align}\label{eq:gen_state_vec}
	\ket\psi &= \left(\begin{matrix}{}
	\alpha_1\\
	\alpha_2\\
	\vdots\\
	\alpha_N
\end{matrix}\right)\nonumber\\
	&= \sum_{n=1}^N \alpha_n \ket{n}.
\end{align}
Here \mbox{$\alpha_n\in\mathbb{C}$} are referred to as the \textit{amplitudes} of the state, which for normalisation satisfy,
\begin{align}\label{eq:state_norm_cond}\index{Normalisation}
\sum_n |\alpha_n|^2 = 1.
\end{align}
Contrast this with the mathematical definition for the normalisation of classical probability distributions,
\begin{align}\index{Normalisation}
\sum_n p_n = 1.
\end{align}
The mathematical objects denoted as $\ket{\cdot}$ are commonly referred to as \textit{kets}\index{Kets}.

In this book we will largely be dealing with qubits\index{Qubits}, the simplest unit of quantum information, which are supported by exactly two basis states, denoted $\ket{0}$ and $\ket{1}$. A qubit can therefore be expressed as,
\begin{align}
\ket\psi &= \left(\begin{matrix}{}
	\alpha \\
	\beta
\end{matrix}\right)\nonumber\\
&= \alpha\ket{0} + \beta\ket{1}.
\end{align}
These are referred to as the \textit{logical basis states}\index{Logical basis states}, which are treated abstractly, detached from a particular physical realisation (e.g by associating them with electron energy levels, photon polarisation, or indeed \textit{any} two-level quantum system), much in the same way that classical bits are treated abstractly as a unit of information, detached from physical reality.

Also mathematically useful is the \textit{dual} of a state vector\index{Dual vector}, which is simply the complex conjugate transpose ($\dag$) of the vector, capturing the same information about the state in conjugate form,
\begin{align}
\bra\psi &= \left(\begin{matrix}{}
	\alpha \\
	\beta
\end{matrix}\right)^\dag\nonumber\\
&=\left(\begin{matrix}{}
	\alpha^*, \beta^*
\end{matrix}\right)\nonumber\\
&= \alpha^*\bra{0} + \beta^*\bra{1}.
\end{align}
The objects $\bra{\cdot}$ are referred to as \textit{bras}\index{Bras}.

@@

For two generic state vectors, supported by the same basis,
\begin{align}
	\ket\psi = \sum_n \alpha_n \ket{n},\nonumber\\
	\ket\phi = \sum_n \beta_n \ket{n},
\end{align}
the \textit{overlap}\index{Overlap} between them is defined as,
\begin{align}
\braket{\psi|\phi} = \sum_n \alpha_n\beta_n^*,
\end{align}
which is mathematically equivalent to their vector dot product\index{Dot product}. The object $\braket{\cdot|\cdot}$ is referred to as a \textit{braket}\footnote{We never quite appreciated the corny sense of humour of the author of the terms \textit{bra}, \textit{ket}, and \textit{braket}. But somehow it stuck.}\index{Braket}. Owing to the cyclic property of the trace operator\index{Trace operator}, it follows that the trace,
\begin{align}
\mathrm{tr}(\ket\psi\bra\phi)=\braket{\psi|\phi},	
\end{align}
is a state overlap. This quantity has the useful intuitive interpretation as the \textit{distinguishability}\index{Distinguishability} between quantum states, which plays an important role in measurement theory and the theory of quantum information. Thus, the normalisation condition from Eq.~(\ref{eq:state_norm_cond}) can equivalently be written,
\begin{align}\index{Normalisation}
\braket{\psi|\psi} = 1.	
\end{align}
Because basis states are orthonormal\index{Orthonormality}, this implies that for basis states $\ket{m}$ and $\ket{n}$ we have the trivial delta function overlap identity,
\begin{align}
	\braket{m|n}=\delta_{m,n}.
\end{align}

@@

More generally, we may also have states of the form,
\begin{align}
\ket\psi_{A,B} =\sum_{m,n} \lambda_{m,n}\ket{m}_A\otimes\ket{n}_B,
\end{align}
where in general $\lambda_{m,n}$ may not be separable as,
\begin{align}
\lambda_{m,n}=\alpha_m\beta_n.
\end{align}
In this case no complete physical description exists for either subsystem in isolation -- they may only be described collectively. Now the state is said to be \textit{entangled}\index{Entanglement}, a uniquely quantum class of physical states with no classical analogue. The simplest and best-known example is the \textit{Bell state}\index{Bell states} on two qubits,
\begin{align}\label{eq:bell_state_vec_def}
\ket\Phi^+_{A,B} &= \frac{1}{\sqrt{2}}\left(\begin{matrix}{}
  1\\
  0\\
  0\\
  1
\end{matrix}\right)\nonumber\\
&= \frac{1}{\sqrt{2}}(\ket{0}_A\ket{0}_B+\ket{1}_A\ket{1}_B).
\end{align}
It's obvious upon inspection that the Bell state cannot be written in separable form,
\begin{align}
	\ket\Phi^+_{A,B} \neq \ket\psi_A \otimes \ket\phi_B,
\end{align}
and is hence entangled.

@@

Thus, in the special case of a qubit, density operators are \mbox{$2\times 2$} complex matrices,
\begin{align}\label{eq:2x2density}
\hat\rho = \left(\begin{matrix}{}
  a & c \\
  c^* & b
\end{matrix}\right),
\end{align}
where diagonal elements are necessarily real, and off-diagonal ones may be complex in general.

The intuitive physical interpretation of density matrices is as follows:
\begin{itemize}
\item Rows and columns enumerate the basis states of the Hilbert space.
\item Diagonal elements represent the classical probabilities\index{Classical probabilities} of the respective basis states.
\item Off-diagonal elements represent \textit{coherences}\index{Coherences} between pairs of basis states, i.e whether they exist as a classical probability distribution (when the coherence is zero) or as a quantum superposition (when non-zero). The $\hat\rho_{m,n}$ matrix element represents the degree of coherence between basis states $\ket{m}$ and $\ket{n}$.
\end{itemize}

Therefore, referring back to the qubit example in Eq.~(\ref{eq:2x2density}), $a$ and $b$ can be regarded as the \textit{classical} probability amplitudes associated with being in the $\ket{0}$ or $\ket{1}$ basis state, while $c$ characterises to what extent it is a classical mixture of these two states (for \mbox{$c=0$}) or a coherent superposition of them (for \mbox{$|c|=ab$}).

@@

For a pure state (a qubit in this example), the density operator takes the form of a vector outer product\index{Outer product},
\begin{align}
\hat\rho &= \ket\psi\bra\psi\nonumber\\
&= \left(\begin{matrix}{}
  |\alpha|^2 & \alpha\beta^* \nonumber\\
  \alpha^*\beta & |\beta|^2
\end{matrix}\right).
\end{align}

@@

The partial trace is a linear operator defined as satisfying the following properties\index{Partial trace!Properties},
\begin{align}
\mathrm{tr}_B(\hat\rho_A\otimes\hat\rho_B) &= \mathrm{tr}(\hat\rho_B) \cdot \hat\rho_A,\nonumber\\
\mathrm{tr}_B(\ket{i}_A\bra{j}_A\otimes\ket{k}_B\bra{l}_B) &= \braket{k|l}\cdot\ket{i}_A\bra{j}_A,\nonumber\\
\mathrm{tr}_B(\hat\rho+\hat\sigma) &= \mathrm{tr}_B(\hat\rho) + \mathrm{tr}_B(\hat\sigma).
\end{align}

@@

We define a measurement \textit{observable}\index{Observable} as an operator $\hat{O}$ acting on a state-space, with eigenvalues $O_i$ over some range of $i$. The eigenvalues $O_i$ denote the allowed values of the measurement outcome for the associated eigenstate, which corresponds to the projected post-measurement state. The statistical expectation value\index{Expectation value} of these measurement outcomes for a given state $\ket\psi$ may now be obtained as the expectation value of the observable,
\begin{align}
	\braket{\hat{O}} = \braket{\psi|\hat{O}|\psi},
\end{align}
and their variance\index{Variance} by,
\begin{align}
	\braket{\hat{O}^2} - \braket{\hat{O}}^2.
\end{align}
Both of these are in the limit of an infinite number of independent trials of the measurement on identical copies of the state.

@@

Next we will apply the so-called Hadamard gate\index{Hadamard gate}, given by the \mbox{$2\times 2$} unitary matrix,
\begin{align}
\hat{H} = \frac{1}{\sqrt{2}}\left(\begin{matrix}{}
  1 & 1 \\
  1 & -1
\end{matrix}\right).
\end{align}

Applying this gate to our input state yields,
\begin{align}
	\ket\psi &= \hat{H}\ket{0}\nonumber\\
	&= \frac{1}{\sqrt{2}}\left(\begin{matrix}{}
  1 & 1 \\
  1 & -1
\end{matrix}\right)\left(\begin{matrix}{}
1\\
0	
\end{matrix}\right)\nonumber\\
&=\frac{1}{\sqrt{2}}\left(\begin{matrix}{}
1\\
1	
\end{matrix}\right)\nonumber\\
&= \frac{1}{\sqrt{2}}(\ket{0}+\ket{1}),
\end{align}
which is an equal superposition of the two logical basis states.

>>>>

\dropcap{T}{he} laws of quantum mechanics are inherently different to those of classical mechanics in virtually every respect, and bring with them their own unique mathematical notation for:
\begin{itemize}
	\item The representation of quantum states.
	\item The evolution of quantum systems.
	\item The measurement of quantum states.
\end{itemize}
We review all of these formalisms in the upcoming sections, and present the necessary mathematical framework to enable following the derivations presented here.

@@

\subsection{Features of quantum mechanics}\index{Quantum mechanics!Features}

Classical and quantum physics are fundamentally different in a number of extremely important ways. These differences are what allow quantum mechanics to enable the entire new field of quantum technologies, most notably quantum computing, which offers massive computational gains over traditional classical computing. The same applies to the novel and groundbreaking promises of the other key quantum technologies, such as quantum cryptography, quantum teleportation, and quantum-enhanced metrology. All of these gains are achieved by exploiting these unique new characteristics offered by quantum systems, that are physically impossible in a classical context -- all technologies must obey the laws of physics, and therefore new laws of physics necessarily imply new technological possibilities.

The key conceptual differences between classical and quantum mechanics are:
\begin{itemize}
	\item Superposition\index{Superpositions}: the ability of quantum systems to be in multiple states simultaneously (simultaneously, not probabilistically!). This arises since quantum states can always be interpreted as waves, which are inherently delocalised and non-particle-like.
	\item Quantum amplitudes: quantum states are not governed by classical probabilities, but by \textit{quantum amplitudes}, which are complex numbers rather than real numbers.\index{Quantum amplitudes}
	\item Entanglement: a type of non-local quantum correlation, stronger than any possible classical correlation.\index{Entanglement}
	\item Unitary evolution: the free evolution of quantum systems is restricted to being described by unitary operators. This implies that all quantum evolution is perfectly time-reversible\index{Time-reversal}, since all unitary operators are invertible.\index{Evolution}
	\item Measurement collapse: the measurement of a quantum system changes its state in general, \textit{collapsing} it to the state corresponding to the measurement outcome\footnote{An open and intuitively confusing conundrum in quantum theory is to resolve the question of what exactly it is that constitutes `measurement' -- the \textit{measurement paradox}\index{Measurement paradox}. Given that all quantum evolution is supposed to be governed by unitary dynamics, whereas measurements are not, at what point in physical reality does an operation transition from one to the other? This is an extremely important open question, particularly from the perspective of quantum foundations and the philosophy of physics. This has led to the emergence of countless theories for the interpretation of quantum mechanics\index{Interpretation of quantum mechanics}, which is an essentially untestable field of research, more akin to fundamentalist religious movements than physical science. However, the resolution is largely unimportant in a practical technological context, where it suffices to define `measurement' as being `when quantum information becomes classical information', e.g once we are able to read it off a measurement device's LCD screen!}.\index{Measurement!Collapse}
	\item Non-determinism: quantum systems exhibit genuine non-determinism, associated with the randomness of the measurement collapse\index{Measurement!Collapse} process\footnote{When we say `random' we \textit{really} mean random, not `quasi-random but actually deterministic'\index{Quasi-randomness}. This true randomness is another of the unique and surprising features of quantum mechanics. Conversely to classical physics, quantum mechanics is inherently and provably non-deterministic! Specifically, it has been proven theoretically \cite{bib:BellNoHiddenVariables} that so-called `hidden variable theories'\index{Hidden variable theories} are incompatible with quantum physics, which rules out any form of quasi-randomness and mandates true randomness.}.\index{Non-determinism}
	\item Exponential complexity: the number of unique parameters required to fully describe a quantum system grows exponentially with its `size' (e.g the number of constituent subsystems, such as quantum particles or qubits). This is in contrast to classical systems whose state complexity grows only linearly with system size, since it's always possible to describe constituent subsystems independently (there is no notion of entanglement). Therefore the number of parameters required to characterise classical systems accumulates additively, whereas for quantum systems it grows multiplicatively.\index{Exponential complexity}
	\item Wave-particle duality: all quantum objects can simultaneously be considered to be particles and waves, where the wave-like and particle-like characteristics express themselves differently depending on the application.\index{Wave-particle duality}
	\item Uncertainty principle: unlike classical systems, where we can essentially know everything about them to arbitrary accuracy, quantum systems necessarily exhibit uncertainties. Heisenberg's uncertainty principle places strict physical bounds on the amount of information that can be known about quantum systems, exhibiting tradeoffs between how much we can know about different parameters simultaneously. The best-known example of this is the position-momentum uncertainty principle, which tells us that the more information we know about an object's position, the less we can simultaneously know about its momentum, and vice versa. It is not possible even in principle to have perfect information about both (this is a physical limitation, not an engineering one!).\index{Uncertainty principle}
\end{itemize}

@@

Mathematically, we can employ the following alternate state vector representations, where the latter is referred to as \textit{Dirac notation} (with $\ket{n}$ denoting the basis states),
\begin{align}\label{eq:gen_state_vec}
	\ket\psi &= \begin{pmatrix}
	\alpha_1\\
	\alpha_2\\
	\vdots\\
	\alpha_N
\end{pmatrix}\nonumber\\
	&= \sum_{n=1}^N \alpha_n \ket{n}.
\end{align}
Here \mbox{$\alpha_n\in\mathbb{C}$} are referred to as the \textit{amplitudes} of the state, which for normalisation satisfy,
\begin{align}\label{eq:state_norm_cond}\index{Normalisation}
\sum_n |\alpha_n|^2 = 1.
\end{align}
Contrast this with the mathematical definition for the normalisation of classical probability distributions,
\begin{align}\index{Normalisation}
\sum_n p_n = 1.
\end{align}
The mathematical objects denoted as $\ket{\cdot}$ are commonly referred to as \textit{kets}\index{Kets}.

In this book we will largely be dealing with qubits\index{Qubits}, the simplest unit of quantum information, which are supported by exactly two basis states, denoted $\ket{0}$ and $\ket{1}$. A qubit can therefore be expressed as,
\begin{align}
\ket\psi &= \begin{pmatrix}
	\alpha \\
	\beta
\end{pmatrix}\nonumber\\
&= \alpha\ket{0} + \beta\ket{1}.
\end{align}
These are referred to as the \textit{logical basis states}\index{Logical basis states}, which are treated abstractly, detached from a particular physical realisation (e.g by associating them with electron energy levels, photon polarisation, or indeed \textit{any} two-level quantum system), much in the same way that classical bits are treated abstractly as a unit of information, detached from physical reality.

Also mathematically useful is the \textit{dual} of a state vector\index{Dual vector}, which is simply the complex conjugate transpose ($\dag$) of the vector, capturing the same information about the state in conjugate form,
\begin{align}
\bra\psi &= \begin{pmatrix}
	\alpha \\
	\beta
\end{pmatrix}^\dag\nonumber\\
&=\begin{pmatrix}
	\alpha^*, \beta^*
\end{pmatrix}\nonumber\\
&= \alpha^*\bra{0} + \beta^*\bra{1}.
\end{align}
The objects $\bra{\cdot}$ are referred to as \textit{bras}\index{Bras}.

@@

For two generic state vectors, supported by the same basis,
\begin{align}
	\ket\psi = \sum_n \alpha_n \ket{n},\nonumber\\
	\ket\phi = \sum_n \beta_n \ket{n},
\end{align}
the \textit{overlap}\index{Overlap} between them is defined as,
\begin{align}
\braket{\psi|\phi} = \sum_n \alpha_n\beta_n^*,
\end{align}
which is mathematically equivalent to their vector dot product\index{Dot product}. The magnitude of the overlap of two quantum states lies in the range,
\begin{align}
	0\leq|\braket{\psi|\phi}|\leq 1,
\end{align}
where \mbox{$|\braket{\psi|\phi}|=0$} indicates that the two states were orthogonal, and \mbox{$|\braket{\psi|\phi}|=1$} implies they are identical (up to an irrelevant global phase\index{Global phase}\footnote{In quantum mechanics, all quantum states are physically invariant under the introduction of a \textit{global} phase-factor, i.e \mbox{$\ket\psi\to e^{i\theta}\ket\psi$}. Thus, global phases may always be safely ignored without changing the underlying physical reality. This most certainly does not apply to \textit{local} phases, which are a part of physical reality, and critically affect interference dynamics.}). The object $\braket{\cdot|\cdot}$ is referred to as a \textit{braket}\footnote{We never quite appreciated the corny sense of humour of the author of the terms \textit{bra}, \textit{ket}, and \textit{braket}. But somehow it stuck.}\index{Braket}. The normalisation condition from Eq.~(\ref{eq:state_norm_cond}) can therefore equivalently be written,
\begin{align}\index{Normalisation}
\braket{\psi|\psi} = 1.	
\end{align}
Owing to the cyclic property of the trace operator\index{Trace operator}, it follows that the trace,
\begin{align}
\mathrm{tr}(\ket\psi\bra\phi)=\braket{\psi|\phi},	
\end{align}
is a state overlap. This quantity has the useful intuitive interpretation as the \textit{distinguishability}\index{Distinguishability} between quantum states, which plays an important role in measurement theory and the theory of quantum information. It is also widely used to characterise the quality of quantum experiments, as a way of comparing quantum states prepared in the lab with the ideal states obtained from the theory.

Because basis states are orthonormal\index{Orthonormality}, this implies that for basis states $\ket{m}$ and $\ket{n}$ we have the trivial delta function overlap identity,
\begin{align}
	\braket{m|n}=\delta_{m,n}.
\end{align}

@@

More generally, we may also have states of the form,
\begin{align}
\ket\psi_{A,B} =\sum_{m,n} \lambda_{m,n}\ket{m}_A\otimes\ket{n}_B,
\end{align}
where in general $\lambda_{m,n}$ may not be separable as,
\begin{align}
\lambda_{m,n}=\alpha_m\beta_n.
\end{align}
In this case no complete physical description exists for either subsystem in isolation -- they may only be described collectively. Now the state is said to be \textit{entangled}\index{Entanglement}, a uniquely quantum class of physical states with no classical analogue. The simplest and best-known example is the \textit{Bell state}\index{Bell states} on two qubits,
\begin{align}\label{eq:bell_state_vec_def}
\ket\Phi^+_{A,B} &= \frac{1}{\sqrt{2}}\begin{pmatrix}
  1\\
  0\\
  0\\
  1
\end{pmatrix}\nonumber\\
&= \frac{1}{\sqrt{2}}(\ket{0}_A\ket{0}_B+\ket{1}_A\ket{1}_B).
\end{align}
It's obvious upon inspection that the Bell state cannot be written in separable form,
\begin{align}
	\ket\Phi^+_{A,B} \neq \ket\psi_A \otimes \ket\phi_B,
\end{align}
and is hence entangled.

@@

Thus, in the special case of a qubit, density operators are \mbox{$2\times 2$} complex matrices,
\begin{align}\label{eq:2x2density}
\hat\rho = \begin{pmatrix}
  a & b \\
  b^* & 1-a
\end{pmatrix},
\end{align}
where diagonal elements are necessarily real, and off-diagonal ones may be complex in general.

The intuitive physical interpretation of density matrices is as follows:
\begin{itemize}
\item Rows and columns enumerate the basis states of the Hilbert space.
\item Diagonal elements represent the classical probabilities\index{Classical probabilities} of the respective basis states.
\item Off-diagonal elements represent \textit{coherences}\index{Coherences} between pairs of basis states, i.e whether they exist as a classical probability distribution (when the coherence is zero) or as a quantum superposition (when non-zero). The $\hat\rho_{m,n}$ matrix element (for \mbox{$m\neq n$}) represents the degree of coherence between basis states $\ket{m}$ and $\ket{n}$.
\end{itemize}

Therefore, referring back to the qubit example in Eq.~(\ref{eq:2x2density}), $a$ and \mbox{$1-a$} can be regarded as the \textit{classical} probability amplitudes associated with being in the $\ket{0}$ or $\ket{1}$ basis state, while $b$ characterises to what extent it is a classical mixture of these two states (for \mbox{$b=0$}) or a coherent superposition of them (for \mbox{$|b|=a(1-a)$}).

@@

For a pure state (a qubit in this example), the density operator takes the form of a vector outer product\index{Outer product},
\begin{align}
\hat\rho &= \ket\psi\bra\psi\nonumber\\
&= \begin{pmatrix}
  |\alpha|^2 & \alpha\beta^* \nonumber\\
  \alpha^*\beta & |\beta|^2
\end{pmatrix}.
\end{align}

@@

The partial trace is a linear operator defined to satisfy the following properties\index{Partial trace!Properties},
\begin{align}
\mathrm{tr}_B(\hat\rho_A\otimes\hat\rho_B) &= \mathrm{tr}(\hat\rho_B) \cdot \hat\rho_A,\nonumber\\
\mathrm{tr}_B(\ket{i}_A\bra{j}_A\otimes\ket{k}_B\bra{l}_B) &= \braket{k|l}\cdot\ket{i}_A\bra{j}_A,\nonumber\\
\mathrm{tr}_B(\hat\rho+\hat\sigma) &= \mathrm{tr}_B(\hat\rho) + \mathrm{tr}_B(\hat\sigma).
\end{align}

@@

We define a measurement \textit{observable}\index{Observable} as an operator $\hat{O}$ acting on a state-space, with eigenvalues $O_i$ over some range of $i$. The eigenvalues $O_i$ denote the allowed values of the measurement outcome for the associated eigenstate, which corresponds to the projected post-measurement state. The statistical expectation value\index{Expectation value} of these measurement outcomes for a given state $\ket\psi$ may now be obtained as the expectation value of the observable,
\begin{align}
	\braket{\hat{O}} = \braket{\psi|\hat{O}|\psi},
\end{align}
and their variance\index{Variance} by,
\begin{align}
	\braket{\hat{O}^2} - \braket{\hat{O}}^2.
\end{align}
Both of these are in the limit of an infinite number of independent trials of the measurement performed on identical copies of the state.

@@

Next we will apply the so-called Hadamard gate\index{Hadamard gate}, given by the \mbox{$2\times 2$} unitary matrix,
\begin{align}
\hat{H} = \frac{1}{\sqrt{2}}\begin{pmatrix}
  1 & 1 \\
  1 & -1
\end{pmatrix}.
\end{align}

Applying this gate to our input state yields,
\begin{align}
	\ket\psi &= \hat{H}\ket{0}\nonumber\\
	&= \frac{1}{\sqrt{2}}\begin{pmatrix}
  1 & 1 \\
  1 & -1
\end{pmatrix}\begin{pmatrix}
1\\
0	
\end{pmatrix}\nonumber\\
&=\frac{1}{\sqrt{2}}\begin{pmatrix}
1\\
1	
\end{pmatrix}\nonumber\\
&= \frac{1}{\sqrt{2}}(\ket{0}+\ket{1}),
\end{align}
which is an equal superposition of the two logical basis states.

### 2018-12-28 Sections/intro_to_QO.tex
Formatting update

These operators satisfy the following basic algebraic properties:
\begin{align}
\hat{a}^\dag\ket{n} &= \sqrt{n+1}\ket{n+1},\nonumber\\
\hat{a}\ket{n} &= \sqrt{n}\ket{n-1},\nonumber\\
\hat{a}\ket{0} &= 0,\nonumber\\
\ket{n} &= \frac{1}{\sqrt{n!}}(\hat{a}^\dag)^n\ket{0}.
\end{align}

The creation and annihilation operators obey the commutation relation\index{Commutation relations},
\begin{align}
	[\hat{a},\hat{a}^\dag] &= 1,
\end{align}
where,
\begin{align}
[\hat{A},\hat{B}]=\hat{A}\hat{B}-\hat{B}\hat{A},
\end{align}
is the \textit{commutator}\index{Commutators} of two operators (operators that commute obviously necessarily exhibit \mbox{$[\hat{A},\hat{B}]=0$}). For multiple modes ($i$ and $j$) the different creation operators commute, as do the respective annihilation operators,
\begin{align}
[\hat{a}^\dag_i,\hat{a}^\dag_j] &= 0,\nonumber\\
[\hat{a}_i,\hat{a}_j] &= 0.
\end{align}
The physical interpretation of this property is that because photons are bosons, they are exchange-symmetric, and thus the physical state is invariant under swapping them.

@@

A particularly ubiquitous state that emerges in CV representations is the \textit{coherent state}\index{Coherent states}, which closely approximates classical laser light\index{Laser},
\begin{align}\index{Coherent states}
\ket{\alpha} &= e^{-\frac{|\alpha|^2}{2}} \sum_{n=0}^\infty \frac{\alpha^n}{\sqrt{n!}} \ket{n},\nonumber\\
\hat{a}\ket{\alpha} &= \alpha\ket{\alpha},
\end{align}
where \mbox{$\alpha\in\mathbb{C}$} is the complex coherent amplitude. Writing the coherent amplitude as,
\begin{align}
	\alpha=re^{i\theta},
\end{align}
$r$ can be interpreted as the strength of the classical field, and $\theta$ as its local phase.

@@

We can graphically represent phase-space on the complex plane, where the two axes denote expectation values of the $\hat{x}$ and $\hat{p}$ operators. Unlike classical phase-space, in quantum phase-space states are `blobs'\index{Blobs} rather than points. The variance of these blobs represents the measurement uncertainty in a given direction, which is strictly non-zero in the quantum world, owing to the Heisenberg uncertainty principle\index{Heisenberg uncertainty principle}.

>>>>

These operators satisfy the following basic algebraic properties:
\begin{align}
\hat{a}^\dag\ket{n} &= \sqrt{n+1}\ket{n+1},\nonumber\\
\hat{a}\ket{n} &= \sqrt{n}\ket{n-1},\nonumber\\
\hat{a}\ket{0} &= 0.
\end{align}
From this, it follows that an arbitrary Fock state can be expressed in terms of creation operators as,
\begin{align}
\ket{n} = \frac{1}{\sqrt{n!}}(\hat{a}^\dag)^n\ket{0}.
\end{align}
The creation and annihilation operators obey the commutation relation\index{Commutation relations},
\begin{align}
	[\hat{a},\hat{a}^\dag] &= 1,
\end{align}
where,
\begin{align}
[\hat{A},\hat{B}]=\hat{A}\hat{B}-\hat{B}\hat{A},
\end{align}
is the \textit{commutator}\index{Commutators} of two operators (operators that commute obviously necessarily exhibit \mbox{$[\hat{A},\hat{B}]=0$}). For multiple modes ($i$ and $j$) the different creation operators commute, as do the respective annihilation operators,
\begin{align}
[\hat{a}^\dag_i,\hat{a}^\dag_j] &= 0,\nonumber\\
[\hat{a}_i,\hat{a}_j] &= 0.
\end{align}
The physical interpretation of this property is that because photons are bosons, they are exchange-symmetric, and thus the physical state is invariant under swapping them.

@@

A particularly ubiquitous state that emerges in CV representations is the \textit{coherent state}\index{Coherent states}, which closely approximates classical laser light\index{Laser},
\begin{align}\index{Coherent states}
\ket{\alpha} = e^{-\frac{|\alpha|^2}{2}} \sum_{n=0}^\infty \frac{\alpha^n}{\sqrt{n!}} \ket{n},
\end{align}
where \mbox{$\alpha\in\mathbb{C}$} is the complex coherent amplitude. Alternately, coherent states can be defined as the eigenstates of the annihilation operator,
\begin{align}
\hat{a}\ket{\alpha} = \alpha\ket{\alpha}.
\end{align}
Writing the coherent amplitude as,
\begin{align}
	\alpha=re^{i\theta},
\end{align}
$r$ can be interpreted as the strength of the classical field, and $\theta$ as its local phase.

@@

We can graphically represent phase-space on the complex plane, where the two axes denote expectation values of the $\hat{x}$ and $\hat{p}$ operators. Unlike classical phase-space, in quantum phase-space states are `blobs'\index{Blobs} rather than points. The variance of these blobs represents the measurement uncertainty in a given direction, which is strictly non-zero in the quantum world, owing to the Heisenberg uncertainty principle\index{Uncertainty principle}.

### 2018-12-28 Sections/mathematical_foundations.tex
Formatting update

This part ought provide the mathematical background necessary to follow the upcoming discussions. Those already familiar with the laws and formulation of quantum mechanics and quantum optics can calmly skip this part.

However, rather than just present the mathematical formulations and expect you to just take the maths as the truth, the light, and the way, we will try as best we can to inject some intuition and meaningful interpretation of the mathematics into our tutorial, so that hopefully the reader is able to walk away with not just the ability to follow basic quantum mechanical derivations, but also grasp the intuition of the underlying physical reality.

>>>>

This part ought provide the mathematical background necessary to follow the upcoming discussions. Those already familiar with the laws and formulation of quantum mechanics and quantum optics can calmly skip this part, without risk of offending the authors\footnote{Except Zixin. She'll probably be offended. \#PC \#SJW.\index{SJW}}.

However, rather than just present the mathematical formulations and expect you to reverently accept the maths as the truth, the light, and the way, we will try as best we can to inject some intuition and meaningful interpretation of the mathematics into our tutorial, so that hopefully the reader is able to walk away with not just the ability to follow basic quantum mechanical derivations, but also grasp the intuition of the underlying physical reality.
